\Large \bf Vigor: Toward Carrier-Grade Software Network Functions

### 2016-10-28 doc/paper/paper.tex
Expand the example with the corectness proof, and more explanations.

\subsection*{Abstract}
The objective of this work is to enable developers to productively write
software-based data plane network functions that simultaneously meet the highest
standards of reliability, security, and performance. We present a verification
approach for formally proving properties of stateful network functions, in a way
that is practical both in terms of developer effort and the performance of the
resulting data plane software. Unlike alternative approaches that require
drastic changes in the way network functions are developed, our approach allows
the use of C/C++ together with popular frameworks like DPDK. To the best of our
knowledge, our work is the first to both be practical and formally verify
stateful network functions. We introduce a new ``speculative'' form of symbolic
execution, which achieves scalability to real-world network functions by
leveraging insights related to the structure and cyclomatic complexity of data
plane software. This technique dovetails with a new library of verified
components that are commonly used in network functions. We illustrate our
approach by developing a NAT box that is verified to never crash, to not leak
memory, be free of several classes of security vulnerabilities, and whose
performance is on par with other NATs. %\footnote{Disclaimer: we use acronym NAT in this paper to refer to a more narrow kind - Dynamic Source NAPT (network address-port translation~\cite{rfc3022}).}. 
We view this as an important milestone on the road to complete verification of all software-based network functions.

@@

To verify that this NF does not crash, one could use the approach of Dobrescu et
al.~\cite{dobrescu2014software} and run exhaustive symbolic execution. The
symbolic execution (SE) technique entails using a symbolic execution engine
(SEE) to explore all paths through the code and check whether any failure
conditions hold --- in our case crash conditions. The inputs to the program are
marked ``symbolic,'' meaning that they subsume all possible input values, and
then the SEE simulates the execution of the program while propagating the
symbolic inputs along the dataflow paths into the program's variables. When
facing a branch (e.g., an \code{if} or a \code{for} loop), the SEE clones the
program state and takes each feasible path separately. Branches impose
constraints on variable values (e.g., an \code{if} statement imposes
restrictions on the values of the condition variable along the \code{else}
branch); the SEE collects these constraints along each path. When evaluating a
branch condition, the SEE asks an SMT constraint solver which of the branches
are feasible, given the constraints collected so far on that path prefix.

@@

{\bf Step 1: Add loop invariants.} Infinite (or very long) loops can be
simplified using loop invariants~\cite[\S~2.1]{cormen2009introduction}. These
are annotations that inform the SEE about invariant properties of the loop, such
as ``no element of \code{packets} between \code{begin} and \code{begin +
  queue_len} has destination port equal to 9'', and are a good match for
NF-style loops.  We added to KLEE support for havoc-ing~\cite{barnett2005boogie}
automatically, to take advantage of such invariants and essentially reduce the
infinite loop to a single abstract iteration that can be reasoned about in one
pass.

This change makes SE feasible, albeit impractical: for \code{CAP=5}, it takes 10
sec to exhaustively SE the code; for \code{CAP=30} it takes 15 min, for
\code{CAP=60} it takes 3 hours, and so on. It takes long due to another classic
SE challenge, ``symbolic
indexing''~\cite{sen2005cute,godefroid2008automated,boonstoppel2008rwset}: when
indexing an array with a symbolic variable (or when dereferencing a symbolic
pointer), the constraints on program state become extremely complex, because the
symbolic pointer or index could be referencing many different possible
locations. The complexity of the constraints chokes the SMT solver, and the SEE
grinds to a halt. In general, SE doesn't work for code with symbolic pointers or
offsets. Alas, stateful C code almost always has some of these.

@@

For our example, the library provides a \code{ring} with a simple interface:
\code{bool ring_empty()} checks whether there is a packet in the ring,
\code{bool ring_full()} checks whether the ring has run out of space, \code{void
  ring_push_back(struct packet* src)} adds packet \code{src} to the ring, and
\code{void ring_pop_front(struct packet* dst)} removes a packet from the ring
and places it in \code{dst}. We then write a symbolic model for the \code{ring}
and provide it to the SEE instead of the actual implementation. The NF now
becomes:

@@

For the symbolic execution and future validation, we need also to customize the
general-purpose packet ring with the following definitions:

@@

/*@
  fixpoint bool user_packet_property(packet p) {
    switch(p) { case packet(port): return port != 9; }
  }
  @*/

@@

@@

Suppose we had an overapproximate model:
\begin{figure}[h!]
\begin{lstlisting}
void ring_pop_front(struct ring* r,
                    struct packet* p) {
  FILL_SYMBOLIC(p, sizeof(struct packet),
                "popped_packet");
}
\end{lstlisting}
  \caption{An extract from an overapproximate ring model.}
  \label{lst:overapproximate}
\end{figure}
Then the \code{ASSERT} from the previous listing would fail.

@@

Here is the example of the proven specification for the function
\code{ring_pop_front}. We do not include the certified implementation of this
specification due to space constraints:

\begin{figure}[h!]
\begin{lstlisting}
void ring_pop_front(struct ring* r,
                    struct packet* p);
/*@ requires ringp(r, ?lst,
                   ?property, ?capacity) &*&
             lst != nil &*&
             packetp(p, _); @*/
/*@ ensures ringp(r, tail(lst),
                  property, capacity) &*&
            packetp(p, head(lst)) &*&
            true == property(head(lst)); @*/
\end{lstlisting}
  \caption{A formal contract for the ring\_pop\_front function.}
  \label{lst:contract}
\end{figure}

Note that \code{property(head(lst));} in the post condition ensures
\code{p->port != 9} property.

@@

To validate the model fit for the particular use case, SEE dumps
all the call sequences (30 in this example) that may happen during the execution.
Vigor checks whether the symbolic model enumerated all the possible
outputs for each sequence.

You can see the example call sequence below. It starts with
\code{loop_invariant_produce} that marks the beginning of a loop iteration. It
continues then with calls to \code{ring_full} \code{ring_empty}, noting that
both return \code{false}. The seqeunce ends with a call to
\code{ring_pop_front}, formally specified above. The last line checks that the
value returned by model (\code{arg2.port != 9}) covers all the cases prescribed
by the specification.

@@

loop_invariant_produce(&(arg1));
/*@ open loop_invariant(_); @*/
bool ret1 = ring_full(arg1);
//@ assume(ret1 == false);
bool ret2 = ring_empty(arg1);
//@ assume(ret2 == false);
//@ close packetp(&(arg2), packet((&(arg2))->port));

@@

Suppose we had an underapproximate model:
\begin{figure}[h!]
\begin{lstlisting}
void ring_pop_front(struct ring* r,
                    struct packet* p) {
  p->port = 0;
}
\end{lstlisting}
  \caption{An underapproximate model fragment.}
  \label{lst:underapproximate}
\end{figure}
then the last line in the call sequence check will look like:
\begin{lstlisting}
/*@ assert(0 == arg2.port); @*/
\end{lstlisting}
And the formal verification will fail, because specification admits a wider
range of possible outputs.

>>>>

\subsection*{Abstract}
The objective of this work is to enable developers to productively write
software-based data plane network functions that simultaneously meet the highest
standards of reliability, security, and performance. We present a verification
approach for formally proving properties of stateful network functions, in a way
that is practical both in terms of developer effort and the performance of the
resulting data plane software. Unlike alternative approaches that require
drastic changes in the way network functions are developed, our approach allows
the use of C/C++ together with popular frameworks like DPDK. To the best of our
knowledge, our work is the first to both be practical and formally verify
stateful network functions. We introduce a new ``speculative'' form of symbolic
execution, which achieves scalability to real-world network functions by
leveraging insights related to the structure and cyclomatic complexity of data
plane software. This technique dovetails with a new library of verified
components that are commonly used in network functions. We illustrate our
approach by developing a NAT box that is verified to never crash, to not leak
memory, be free of several classes of security vulnerabilities, and whose
performance is on par with other NATs.
% \footnote{Disclaimer: we use acronym NAT in this paper to refer to a more
% narrow kind - Dynamic Source NAPT (network address-port
% translation~\cite{rfc3022}).}.
We view this as an important milestone on the road to complete verification of all software-based network functions.

@@

To verify that this NF does not crash, one could use the approach of Dobrescu et
al.~\cite{dobrescu2014software} and run exhaustive symbolic execution. The
symbolic execution (SE) technique entails using a symbolic execution engine
(SEE) to explore all paths through the code and check whether any failure
conditions hold --- in our case crash conditions. The inputs to the program are
marked ``symbolic,'' meaning that they subsume all possible input values, and
then the SEE simulates the execution of the program while propagating the
symbolic inputs along the dataflow paths into the program's variables. When
facing a branch (e.g., an \code{if} or a \code{while} loop), the SEE clones the
program state and takes each feasible path separately. Branches impose
constraints on variable values (e.g., an \code{if} statement imposes
restrictions on the values of the condition variable along the \code{else}
branch); the SEE collects these constraints along each path. When evaluating a
branch condition, the SEE asks an SMT constraint solver which of the branches
are feasible, given the constraints collected so far on that path prefix.

@@

{\bf Step 1: Add loop invariants.} Infinite (or very long) loops can be
simplified using loop invariants~\cite[\S~2.1]{cormen2009introduction}. These
are annotations that inform the SEE about invariant properties of the loop, such
as ``no element of \code{packets} between \code{begin} and \code{begin +
  queue_len} has \code{port == 9}'', and are a good match for NF-style loops. We
added to KLEE support for havoc-ing~\cite{barnett2005boogie} automatically
(replace all modified variables with unconstrained symbols), to take advantage
of such invariants and essentially reduce the infinite loop to a single abstract
iteration that allows reasoning in one pass.

This change makes SE feasible, albeit impractical: for \code{CAP=5}, it takes 10
sec to exhaustively SE the code; for \code{CAP=30} it takes 15 min, for
\code{CAP=60} it takes 3 hours, and so on. It takes so long due to another classic
SE challenge, ``symbolic
indexing''~\cite{sen2005cute,godefroid2008automated,boonstoppel2008rwset}: when
indexing an array with a symbolic variable (or when dereferencing a symbolic
pointer), the constraints on program state become extremely complex, because the
symbolic pointer or index could be referencing many different possible
locations. The complexity of the constraints chokes the SMT solver, and the SEE
grinds to a halt. In general, SE doesn't work for code with symbolic pointers or
offsets. Alas, stateful C code almost always has some of these.

@@

For our example, the library provides a packet ring with a simple interface:
\code{bool ring_empty(ring*)} checks whether there is a packet in the ring,
\code{bool ring_full(ring*)} checks whether the ring has run out of space, \code{void
  ring_push_back(ring*, packet* src)} adds packet \code{src} to the ring, and
\code{void ring_pop_front(ring*, packet* dst)} removes a packet from the ring
and places it in \code{dst}. We then write a symbolic model for the \code{ring}
and provide it to the SEE instead of the actual implementation. The NF now
becomes:

@@

For the symbolic execution and future validation, we need also to customize the
general-purpose packet ring library with the following definitions:

@@

/*@
  fixpoint bool
  user_property(packet p) {
    switch(p) { case packet(port):
                  return port != 9; }
  }
  @*/

@@

Currently, we have to duplicate the packet condition in two languages (formal
logic, and C) due to the limitations of our prototype. This is the
application-specific part of the loop invariant given before, that we plug-in in
order for Vigor to be able to verify the app.

@@

Suppose we had an overapproximate model:
\begin{figure}[h!]
\begin{lstlisting}
void ring_pop_front(struct ring* r,
                    struct packet* p) {
  FILL_SYMBOLIC(p, sizeof(struct packet),
                "popped_packet");
  // No constraints on the packet port here.
}
\end{lstlisting}
  \caption{An extract from an overapproximate ring model.}
  \label{lst:overapproximate}
\end{figure}
Then the \code{VIGOR_CHECK} from the previous listing would fail.

@@

Here is the example of the proven specification for the function
\code{ring_pop_front}:

\begin{figure}[h!]
\begin{lstlisting}
void ring_pop_front(struct ring* r,
                    struct packet* p)
/*@ requires ringp(r, ?user_property,
                   ?lst, ?cap) &*&
             lst != nil &*& packetp(p, _); @*/
/*@ ensures ringp(r, user_property,
                  tail(lst), cap) &*&
            packetp(p, head(lst)) &*&
            true == user_property(head(lst)); @*/
{
  //@ extract_first(r);
  struct packet* src_pkt = r->array + r->begin;
  p->port = src_pkt->port;
  r->len = r->len - 1;
  r->begin = r->begin + 1;
  if (r->cap <= r->begin) {
    r->begin = 0;
    //@ stitch_with_empty_overflow(r);
  } else {
    //@ stitch_with_empty(r);
  }
}
\end{lstlisting}
  \caption{A formal contract and verified implementation of the \code{ring_pop_front} function.}
  \label{lst:contract}
\end{figure}

This C code is intermingled with the correctness proof annotations that help the
formal verification engine (VeriFast) with small steps to transform the initial
state, given in \code{requires} to the final state, specified in \code{ensures}.
After proving the correctness of this transformation, the formal verification
engine will use \code{requires/ensures} formulas as pre- and post- conditions
for \code{ring_pop_front} contract. The trigraph \code{&*&} separates different
statements that hold together. For example

\begin{lstlisting}
ringp(r, user_property, tail(lst), cap) &*&
packetp(p, head(lst)) &*&
true == user_property(head(lst))};
\end{lstlisting}
means:
\begin{itemize}
\item \code{r} points to a healthy ring of capacity \code{cap}, that contains
  \code{tail(lst)} packets, that satisfy the \code{user_property}.
\item \code{p} points to a packet \code{head(lst)}.
\item \code{head(lst)} satisfies the \code{user_property}.
\end{itemize}

Note that \code{user_property(head(lst));} in the post condition ensures
\code{p->port != 9} property.

@@

To validate the model fit for the particular use case, SEE dumps all the call
sequences (30 in this example) extracted from the execution paths. Vigor
checks whether the symbolic model has enumerated all the possible outputs for
each call sequence.

You can see the example call sequence below. It starts with
\code{loop_invariant_produce} that marks the beginning of a loop iteration. It
continues then with calls to \code{ring_full} \code{ring_empty}, noting (with
\code{assume}) that both return \code{false}. The seqeunce ends with a call to
\code{ring_pop_front}, formally specified above. The last line checks that the
value returned by model (\code{arg2.port != 9}) covers all the cases prescribed
by the specification.

@@

loop_invariant_produce(&(arg1));
/*@ open loop_invariant(_); @*/
bool ret1 = ring_full(arg1);
//@ assume(ret1 == false);
bool ret2 = ring_empty(arg1);
//@ assume(ret2 == false);
/*@ close packetp(&(arg2),
                  packet((&(arg2))->port)); @*/

@@

Formal verification engine then checks this sequence using the contracts
like in the Figure~\ref{lst:contract}, and ensures the abscence of misuse. It then checks
that the model output covers all the possible outputs specified in the contract
for the tip function (see~\S\ref{sec:model-validation-criteria}).
\begin{figure}[h!]
\begin{lstlisting}
void ring_pop_front(struct ring* r,
                    struct packet* p) {
  p->port = 0;
}
\end{lstlisting}
  \caption{An underapproximate model fragment.}
  \label{lst:underapproximate}
\end{figure}

Suppose we had an underapproximate model: Figure~\ref{lst:underapproximate}.
Then the last line in the call sequence check will look like:
\begin{lstlisting}
/*@ assert(0 == arg2.port); @*/
\end{lstlisting}
And the formal verification will fail, because the contract
(Figure~\ref{lst:contract}) specifies a wider range of possible outputs.
